04_DESeq_gon

### 2019-04-06 analysis/04_DESeq2_gon.Rmd
deseq gonads

```{r setup}
library(tidyverse)
library(DESeq2)
library(cowplot)
library(RColorBrewer)
library(pheatmap)

@@

```{r colDataWrangle}
# making new groups
colData$group <- NULL
colData$tempgroup <- ifelse(colData$treatment == "bldg", "bldg",
                  ifelse(colData$treatment == "control", "control",
                   ifelse(colData$treatment == "hatch", "hatch",
                    ifelse(grepl("inc", colData$treatment), "inc",
                     ifelse(colData$treatment == "lay", "lay", "nestl")))))
colData$tempgroup <- as.factor(colData$tempgroup)

colData$group <- paste(colData$sex, colData$tempgroup, sep = "")
colData$group <- as.factor(colData$group)
str(colData$group)

colData$treatment <- factor(colData$treatment, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))
```

```{r filter}
#select samples of interest from colData
colData <- colData %>%
  dplyr::filter(grepl('gonad', tissue)) %>%
  #dplyr::filter(treatment != "control") %>%
  dplyr::filter(sex == "female") %>%
  droplevels()
row.names(colData) <- colData$V1

@@

#select samples of interest from countData
savecols <- as.character(colData$V1) 
savecols <- as.vector(savecols) 
countData <- countData %>% dplyr::select(one_of(savecols))

@@

```{r res}
levels(colData$treatment)
summary(results(dds, contrast=c("treatment",'control', 'bldg'))) 
```

```{res2}
summary(results(dds, contrast=c("treatment",'control', 'lay')))     
summary(results(dds, contrast=c("treatment",'control', 'inc.d3')))  
summary(results(dds, contrast=c("treatment",'control', 'inc.d9')))   
summary(results(dds, contrast=c("treatment",'control', 'inc.d17')))  
summary(results(dds, contrast=c("treatment",'control', 'hatch')))    
summary(results(dds, contrast=c("treatment",'control', 'n5')))      
summary(results(dds, contrast=c("treatment",'control', 'n9')))

summary(results(dds, contrast=c("treatment",'bldg', 'lay')))      
summary(results(dds, contrast=c("treatment",'bldg', 'inc.d3')))   
summary(results(dds, contrast=c("treatment",'bldg', 'inc.d9')))   
summary(results(dds, contrast=c("treatment",'bldg', 'inc.d17')))  
summary(results(dds, contrast=c("treatment",'bldg', 'hatch')))   
summary(results(dds, contrast=c("treatment",'bldg', 'n5')))      
summary(results(dds, contrast=c("treatment",'bldg', 'n9')))

#summary(results(dds, contrast=c("treatment",'lay', 'bldg')))     
summary(results(dds, contrast=c("treatment",'lay', 'inc.d3')))    
summary(results(dds, contrast=c("treatment",'lay', 'inc.d9')))   
summary(results(dds, contrast=c("treatment",'lay', 'inc.d17')))  
summary(results(dds, contrast=c("treatment",'lay', 'hatch')))    
summary(results(dds, contrast=c("treatment",'lay', 'n5')))       
summary(results(dds, contrast=c("treatment",'lay', 'n9')))

#summary(results(dds, contrast=c("treatment",'inc.d3', 'bldg')))     
#summary(results(dds, contrast=c("treatment",'inc.d3', 'lay')))       
summary(results(dds, contrast=c("treatment",'inc.d3', 'inc.d9')))   
summary(results(dds, contrast=c("treatment",'inc.d3', 'inc.d17')))   
summary(results(dds, contrast=c("treatment",'inc.d3', 'lay')))      
summary(results(dds, contrast=c("treatment",'inc.d3', 'n5')))        
summary(results(dds, contrast=c("treatment",'inc.d3', 'n9')))

summary(results(dds, contrast=c("treatment",'inc.d9', 'inc.d17')))   
summary(results(dds, contrast=c("treatment",'inc.d9', 'lay')))      
summary(results(dds, contrast=c("treatment",'inc.d9', 'n5')))       
summary(results(dds, contrast=c("treatment",'inc.d9', 'n9')))

summary(results(dds, contrast=c("treatment",'inc.d17', 'lay')))  
summary(results(dds, contrast=c("treatment",'inc.d17', 'n5')))   
summary(results(dds, contrast=c("treatment",'inc.d17', 'n9')))

#summary(results(dds, contrast=c("treatment",'hatch', 'bldg')))      
#summary(results(dds, contrast=c("treatment",'hatch', 'lay')))       
#summary(results(dds, contrast=c("treatment",'hatch', 'inc.d3')))    
#summary(results(dds, contrast=c("treatment",'hatch', 'inc.d9')))    
summary(results(dds, contrast=c("treatment",'hatch', 'inc.d17')))  
summary(results(dds, contrast=c("treatment",'hatch', 'n5')))       
summary(results(dds, contrast=c("treatment",'hatch', 'n9')))

#summary(results(dds, contrast=c("treatment",'n5', 'bldg')))      
#summary(results(dds, contrast=c("treatment",'n5', 'lay')))       
#summary(results(dds, contrast=c("treatment",'n5', 'inc.d3')))  
#summary(results(dds, contrast=c("treatment",'n5', 'inc.d9')))   
#summary(results(dds, contrast=c("treatment",'n5', 'inc.d17')))  
summary(results(dds, contrast=c("treatment",'n5', 'hatch')))   
summary(results(dds, contrast=c("treatment",'n5', 'n9')))

#summary(results(dds, contrast=c("treatment",'n9', 'bldg')))     
#summary(results(dds, contrast=c("treatment",'n9', 'lay')))      
#summary(results(dds, contrast=c("treatment",'n9', 'inc.d3')))   
#summary(results(dds, contrast=c("treatment",'n9', 'inc.d9')))   
#summary(results(dds, contrast=c("treatment",'n9', 'inc.d17')))  
#summary(results(dds, contrast=c("treatment",'n9', 'hatch')))    
summary(results(dds, contrast=c("treatment",'n9', 'n5')))     
```

```{r totalDEGs}
levels(colData$treatment)

totalDEGs(c("treatment", 'bldg','lay'))
totalDEGs(c("treatment", 'bldg','inc.d3'))
totalDEGs(c("treatment", 'bldg','inc.d9'))
totalDEGs(c("treatment", 'bldg','inc.d17'))
totalDEGs(c("treatment", 'bldg','hatch'))
totalDEGs(c("treatment", 'bldg','n5'))
totalDEGs(c("treatment", 'bldg','n9'))

#totalDEGs(c("treatment", 'lay','bldg'))
totalDEGs(c("treatment", 'lay','inc.d3'))
totalDEGs(c("treatment", 'lay','inc.d9'))
totalDEGs(c("treatment", 'lay','inc.d17'))
totalDEGs(c("treatment", 'lay','hatch'))
totalDEGs(c("treatment", 'lay','n5'))
totalDEGs(c("treatment", 'lay','n9'))

#totalDEGs(c("treatment", 'inc.d3','bldg'))
#totalDEGs(c("treatment", 'inc.d3','lay'))
totalDEGs(c("treatment", 'inc.d3','inc.d9'))
totalDEGs(c("treatment", 'inc.d3','inc.d17'))
totalDEGs(c("treatment", 'inc.d3','hatch'))
totalDEGs(c("treatment", 'inc.d3','n5'))
totalDEGs(c("treatment", 'inc.d3','n9'))

#totalDEGs(c("treatment", 'inc.d9','bldg'))
#totalDEGs(c("treatment", 'inc.d9','lay'))
#totalDEGs(c("treatment", 'inc.d9','inc.d3'))
totalDEGs(c("treatment", 'inc.d9','inc.d17'))
totalDEGs(c("treatment", 'inc.d9','hatch'))
totalDEGs(c("treatment", 'inc.d9','n5'))
totalDEGs(c("treatment", 'inc.d9','n9'))

#totalDEGs(c("treatment", 'inc.d17','bldg'))
#totalDEGs(c("treatment", 'inc.d17','lay'))
#totalDEGs(c("treatment", 'inc.d17','inc.d3'))
#totalDEGs(c("treatment", 'inc.d17','inc.d9'))
totalDEGs(c("treatment", 'inc.d17','hatch'))
totalDEGs(c("treatment", 'inc.d17','n5'))
totalDEGs(c("treatment", 'inc.d17','n9'))

#totalDEGs(c("treatment", 'hatch','bldg'))
#totalDEGs(c("treatment", 'hatch','lay'))
#totalDEGs(c("treatment", 'hatch','inc.d3'))
#totalDEGs(c("treatment", 'hatch','inc.d9'))
totalDEGs(c("treatment", 'hatch','inc.d17'))
totalDEGs(c("treatment", 'hatch','n5'))
totalDEGs(c("treatment", 'hatch','n9'))

#totalDEGs(c("treatment", 'n5','bldg'))
#totalDEGs(c("treatment", 'n5','lay'))
#totalDEGs(c("treatment", 'n5','inc.d3'))
#totalDEGs(c("treatment", 'n5','inc.d9'))
#totalDEGs(c("treatment", 'n5','inc.d17'))
totalDEGs(c("treatment", 'n5','hatch'))
totalDEGs(c("treatment", 'n5','n9'))

#totalDEGs(c("treatment", 'n9','bldg'))
#totalDEGs(c("treatment", 'n9','lay'))
#totalDEGs(c("treatment", 'n9','inc.d3'))
#totalDEGs(c("treatment", 'n9','inc.d9'))
#totalDEGs(c("treatment", 'n9','inc.d17'))
#totalDEGs(c("treatment", 'n9','hatch'))
totalDEGs(c("treatment", 'n9','n5'))
```

```{r PCA, message=FALSE, warning=FALSE}
levels(colData$treatment)

# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(vsd, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

@@

ggplot(pcadata, aes(PC2, PC3,color = treatment)) + 
  geom_point(size = 2, alpha = 1) +
  stat_ellipse(type = "t") +
  xlab(paste0("PC2: ", percentVar[2],"% variance")) +
  ylab(paste0("PC3: ", percentVar[3],"% variance")) +
  theme_cowplot(font_size = 8, line_size = 0.25)

@@

plotPCA(vsd, intgroup=c("treatment"))
```

PCA statistics

```{r pcastats}
summary(aov(PC1 ~ treatment, data=pcadata)) 
TukeyHSD(aov(PC1 ~ treatment, data=pcadata), which = "treatment")

@@

@@

>>>>

```{r setup, message=F, comment=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(DESeq2)
library(cowplot)
library(RColorBrewer)
library(pheatmap)
library(kableExtra)
library(viridis)

@@

```{r factors}
colData$treatment <- factor(colData$treatment, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))
```

```{r filter}
colData <- colData %>%
  dplyr::filter(grepl('gonad', tissue)) %>%
  dplyr::filter(sex == "female") %>%
  droplevels()
row.names(colData) <- colData$V1

@@

savecols <- as.character(colData$V1) 
savecols <- as.vector(savecols) 
countData <- countData %>% dplyr::select(one_of(savecols))

@@

```{r resloop}
#create list of groups
a <- levels(colData$treatment)
b <- levels(colData$treatment)

# slim for testing
#a <- c("n9", "bldg" , "lay" )
#b <- c("n9", "bldg" , "lay" )

# comapre all contrasts, save to datafrmes
dat=data.frame()
for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = "") #assigns usique rownames
      dat[k,1]<-i               
      dat[k,2]<-j
      dat[k,3]<- numDEGs(i,j) #caluculates number of DEGs
    }
  }
}

head(dat)
```

```{r restable}
# widen data to create table of degs
rownames(dat) <- NULL #remove row names
data_wide <- spread(dat, V2, V3)
data_wide
kable(data_wide)

dat$V1 <- factor(dat$V1, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))
dat$V2 <- factor(dat$V2, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))

ggplot(dat, aes(V1, V2)) +
  geom_tile(aes(fill = V3)) +
  scale_fill_viridis(na.value="#FFFFFF00") + 
  xlab("Treatment") + ylab("Treatment") +
  labs(fill = "# of DEGs")
```

```{r PCA, message=FALSE, warning=FALSE}
# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(vsd, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

@@

summary(aov(PC1 ~ treatment, data=pcadata)) 
TukeyHSD(aov(PC1 ~ treatment, data=pcadata), which = "treatment")

@@

@@

Heatmaps

@@

# Males only

```{r readMale}
# import "colData" which contains sample information and "countData" which contains read counts
colData <- read.csv("../results/00_colData_characterization.csv", header = T, row.names = 1)
countData <- read.csv("../results/00_countData_characterization.csv", header = T, row.names = 1)
geneinfo <- read.csv("../results/00_geneinfo.csv", row.names = 1)
```

```{r factorsMale}
colData$treatment <- factor(colData$treatment, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))
```

```{r filterMale}
colData <- colData %>%
  dplyr::filter(grepl('gonad', tissue)) %>%
  dplyr::filter(sex == "male") %>%
  droplevels()
row.names(colData) <- colData$V1

savecols <- as.character(colData$V1) 
savecols <- as.vector(savecols) 
countData <- countData %>% dplyr::select(one_of(savecols))

```{r deseq2male}
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ treatment )
dds <- dds[ rowSums(counts(dds)) > 2, ] ## pre-filter genes 
dds <- DESeq(dds) # Differential expression analysis
vsd <- vst(dds, blind=FALSE) # variance stabilized 
```

```{r resloopMale}
#create list of groups
a <- levels(colData$treatment)
b <- levels(colData$treatment)

# slim for testing
#a <- c("n9", "bldg" , "lay" )
#b <- c("n9", "bldg" , "lay" )

# comapre all contrasts, save to datafrmes
dat=data.frame()
for (i in a){
  for (j in b){
    if (i != j) {
      k <- paste(i,j, sep = "") #assigns usique rownames
      dat[k,1]<-i               
      dat[k,2]<-j
      dat[k,3]<- numDEGs(i,j) #caluculates number of DEGs
    }
  }
}

head(dat)
```

```{r restableMale}
# widen data to create table of degs
rownames(dat) <- NULL #remove row names
data_wide <- spread(dat, V2, V3)
data_wide
kable(data_wide)

dat$V1 <- factor(dat$V1, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))
dat$V2 <- factor(dat$V2, levels = 
                              c("control", "bldg", "lay", "inc.d3", "inc.d9", 
                                "inc.d17", "hatch", "n5", "n9"))

ggplot(dat, aes(V1, V2)) +
  geom_tile(aes(fill = V3)) +
  scale_fill_viridis(na.value="#FFFFFF00") + 
  xlab("Treatment") + ylab("Treatment") +
  labs(fill = "# of DEGs")
```
